---
title: "CRF and cancer in women - ActToPrevent"
format: html
---

Data for writing the article: 
“The association between cardiorespiratory fitness and incidence and mortality in common cancer forms in women”
Project: ActToPrevent
Aim: The overall aim is to investigate the association between cardiorespiratory fitness (CRF) and incidence and mortality in common cancer forms in women.
-	What is the association between CRF and incidence of- and death from breast, colon, and lung cancer in adults? 
-	Does age (at time of HPA test, and at age of diagnosis) moderate the associations at any time point? 


Data:
Inclusion
•	Women only
•	Full data for all covariates
•	18 and older

Covariates at HPA
•	Age
•	Educational level
•	BMI
•	Smoking
•	Alcohol
•	Diet

Covariates from national register
•	Co-morbidity; Patientregistret) presence of any disease/disorder in the following categories (range 0-4) 
o	Hypertension (ICD-10 code I10-I13)
o	Cardiovascular disease (I20-I25, I60-I69)
o	Diabetes (E10-E14)
o	Other cancers (C00-99)



```{r}
library(tidyverse)
```


# Data
```{r}
df <- read_csv("../../Data/data2020/rensad data 2020_scb_dor_patreg_fk/data/hbt_scb_cleaned_data_all_longformat_2021-05-21.csv")


pr <- read_csv("../../Data/data2020/rensad data 2020_scb_dor_patreg_fk/data/patregdata_with_translated_icd_codes_2021-08-23.csv")


dr <- read.csv("../../Data/data2020/rensad data 2020_scb_dor_patreg_fk/data/dor_dates_cleaned_2021-07-01.csv")
```



# selecting variables
```{r}
df1 <- df %>% 
  select(
    LopNr,
    Performed,
    YearOfBirth,
    Gender,
    Age,
    Astrand_rel_VO2,
    Astrand_MaxVO2,
    HeightCM,
    WeightKG,
    BMI,
    SUN2000Niva_Old,
    ExerciseAnswer,
    TobaccoSmoking,
    AlcoholFrequency,
    AlcoholStandardDrinks,
    AlcoholTotalScore,
    AlcoholStandardFrequency,
    Diet,
    WaistCircumference
  ) %>%
  drop_na(
    LopNr,
    Performed,
    YearOfBirth,
    Gender,
    Age,
    Astrand_rel_VO2,
    Astrand_MaxVO2,
    HeightCM,
    WeightKG,
    BMI,
    SUN2000Niva_Old,
    ExerciseAnswer,
    TobaccoSmoking,
    AlcoholFrequency,
    AlcoholStandardDrinks,
    AlcoholTotalScore,
    AlcoholStandardFrequency,
    Diet
  ) %>% 
  arrange(LopNr,Performed) %>%
  add_count(LopNr) %>% 
  filter(n == 1)
```




ICD-code
Breast cancer	C50
Colon cancer	C18
Lung cancer	C34


Från Patientregistret
Incidence
hdia första tillfället efter genomförd HPB + gärna variabel med risktid beräknad (dvs tid mellan HPB och insjuknande alt död alt uppföljningsdatumet 2019-12-31). Ha gärna kvar en variabel med vilken kod som bidragit med vilket incident fall (om man vill göra analys per subdiagnos).
Age at diagnosis ELLER year of diagnosis and we work out age of diagnosis from their year of birth


Co-morbiditet inkl ICD-10 kod (binära variabler för de 4 kategorierna a-d), whenever: 

a) Hypertension (ICD-10 code I10-I13)
b) Cardiovascular disease (I20-I25, I60-I69)
c) Diabetes (E10-E14)
d) Andra cancers (C, förutom C18, C34, C50)

Exkluderingsvariabel inför analys
Skapa en variabel som indikerar individer med samma ICD-10 koder som ovan innan HPBn (behöver ingen risktid). Ska exkluderas sen, men behöver ha med i framställan av sample size.

Från dödsorsaksregistret

Mortality
Död med diagnoskoderna ovan (hdia) med risktid (tid mellan HPB och död alt uppföljningsdatumet 2019-12-31).




```{r}
df2 <- df1 |> 
  
  # patientreg 1, incident cases
  left_join(
    
    pr %>%   
        separate(hdia, into = c("letter", "number"), sep = 1, remove = FALSE) %>% 
      filter(hdia == "C50"| hdia == "C18"| hdia == "C34") |> 
      mutate(Cancer = if_else(number>=0,1,0)) |>  
      select(LopNr, hdia, Cancer, indatum, utdatum) |> 
          group_by(LopNr) %>% 
  arrange(LopNr, indatum) %>% 
        mutate(row = row_number()) %>% 
  filter(row == min(row))  |> 
    select(!row)
  ) |> 
        mutate(C_BeforeOrAfter_HPB = if_else(indatum > Performed, "after", "before"),
               exclude = if_else(C_BeforeOrAfter_HPB == "before",1,0),
               incident_after_HPB = if_else(C_BeforeOrAfter_HPB == "after",1,0) ) |> 
  
  
  # patientreg 2, co-morbidity
  left_join(
    pr %>%   
      separate(hdia, into = c("letter", "number"), sep = 1, remove = FALSE) %>% 
      mutate(comorbidity = case_when(letter == "I" &
                                       between(number, 10,13) ~ "Hypertension",
                                     letter == "I" &
                                       between(number, 20,25) |
                                       letter == "I" &
                                       between(number, 60,69)~ "CVD",
                                     letter == "E" &
                                       between(number, 10,14) ~ "Diabetes",
                                     letter == "C" &
                                       number != 18  &
                                       letter == "C" &
                                       number != 34    &
                                       letter == "C" &
                                       number != 50   ~ "Cancer_other"
                                     
      ),
      comorbidity_indatum=indatum 
      ) |> 
      select(LopNr, comorbidity, comorbidity_indatum) |>   
      drop_na(comorbidity) |>
      group_by(LopNr, comorbidity) %>% 
      arrange(LopNr, comorbidity_indatum) %>% 
      mutate(row = row_number()) %>% 
      filter(row == min(row)) |> 
      ungroup() |> 
      pivot_wider(names_from = comorbidity, values_from = comorbidity_indatum) |> 
      rename("comorbidity" = row), 
    by = c("LopNr" = "LopNr")
    ) |> 
  
  # Dödsregistret  
  left_join(
    dr %>% select(LopNr, AterPNr, DODSDAT, ULORSAK) |> 
      separate(ULORSAK, into = c("letter", "number"), sep = c(1,3), remove = FALSE) |> 
      mutate(C_DOD = case_when( str_detect( ULORSAK,  "C50|C18|C34") ~ 1),
      DOD = if_else(number >=0,1,0) 
      ) |> 
      select(LopNr, DODSDAT, DOD, C_DOD, ULORSAK, letter,number) 
    ) |> 
  mutate(
    comorbidity=as.numeric(comorbidity),
    C_DOD=as.numeric(C_DOD),
    DOD=as.numeric(DOD),
    Cancer=as.numeric(Cancer),
    comorbidity = if_else(is.na(comorbidity),0, comorbidity),
    C_DOD = if_else(is.na(C_DOD),0, C_DOD),
    DOD = if_else(is.na(DOD),0, DOD),
    Cancer = if_else(is.na(Cancer),0, Cancer),
    incident_after_HPB = if_else(is.na(incident_after_HPB),0, incident_after_HPB),
    exclude = if_else(is.na(exclude),0, exclude)
  ) 


```



```{r}
df_final <-
  df2 %>% ungroup() %>% 
  mutate(
   # across(c(pat_hdia_before_HPB, pat_after_HPB_hypertension:pat_n_ater, pat_incident_1, pat_incident_2), replace_na, 0),
    # patientreg risktid 1
    Cancer = if_else(is.na(Cancer), as.Date("2019-12-31"), as.Date(indatum)),
   Cancer = if_else(Cancer > DODSDAT & !is.na(DODSDAT), as.Date(DODSDAT), Cancer),
   pat_time = as.numeric(Cancer - as.Date(Performed))
  )


```

```{r}

# df_final |> select(Performed, pat_time, Cancer, incident, exclude, C_BeforeOrAfter_HPB) |> view()

write_csv(df_final, "data/CRFcancerWomen_2023-03-23.csv")
```

